---
title: "Overview statistic tests PB"
author: "Jonas van Nijnatten"
date: "21 november 2018"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_float: true  
    toc_depth: 4
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: '2'
header-includes:
- \usepackage{geometry}
- \geometry{a4paper, portrait, margin=.75in}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


contact: J.J.vanNijnatten@uva.nl

broncode: https://github.com/jonasvannijnatten/R_Data_Visualization  

***

\newpage

#   Software  
## Versies  
software versions used for this tutorial:  
 - `r R.Version()$version.string`  
 - car-package version: `r packageVersion("car")` (`r packageDate("car")`)  
 - ez-package version:  `r packageVersion("ez")` (`r packageDate("ez")`)
 
##  Installatie
Benodigde packages downloaden & installeren:
```{r, eval = FALSE, echo=TRUE, results='hide', message=FALSE}
install.packages(pkgs="kableExtra",   repos="https://www.freestatistics.org/cran/")
install.packages(pkgs="car",   repos="https://www.freestatistics.org/cran/") 
install.packages(pkgs="ez",    repos="https://www.freestatistics.org/cran/")
```

Benodigde packages activeren:
```{r, eval=TRUE, message=FALSE}
library(package=kableExtra)
library(package=car)
library(package=ez)
```

***

\newpage

#  T-test

## Independent samples T-test
<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r}
# generate data
N = 40
data.long = data.frame(
  ID = 1:N,
  condition = rep(x = c("A","B"), each = N),
  score = c(rnorm(n = N, mean = 25, sd = 6.5), rnorm(n = N, mean = 35, sd = 6.5))
)
```
</details>

```{r,echo=FALSE}
kable(data.long[c(1:5,41:45),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Test for assumption of normality
```{r}
by(data = data.long$score, INDICES = data.long$condition, FUN = shapiro.test)
```
Test for equality of variances
```{r}
leveneTest(y = data.long$score, group = data.long$condition)
```
**T-test**
```{r}
t.test(formula = score ~ condition, data = data.long, paired=FALSE, alternative="two.sided", var.equal=TRUE)
```

## Paired samples T-test
<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r}
# generate data
N = 30
data.long = data.frame(
  ID = rep(1:N,2),
  condition = rep(x = c("A","B"), each = N),
  score = c(rnorm(n = N, mean = 25, sd = 6.5), rnorm(n = N, mean = 35, sd = 6.5))
)
```
</details>

```{r,echo=FALSE}
kable(data.long[c(1:5,31:35),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Test for assumption of normality
```{r}
# calculate difference scores
diffScore = data.long$score[data.long$condition=="A"] - data.long$score[data.long$condition=="B"]
shapiro.test(diffScore)
```

**T-test**
```{r}
t.test(formula = score ~ condition, data = data.long, paired=TRUE, alternative="two.sided", var.equal=TRUE)
```

***
\newpage

#  Correlation

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary> 
```{r}
# generate data
set.seed(05)
nrobs = 100
experience = rnorm(n = nrobs, mean = 15, sd = 3)
salary     = 10000 + ( 5 * experience ) + rnorm(n = nrobs, mean = 0, sd = 100)
data.long = data.frame(experience, salary)
# calculate correlation coefficient r
corr_coef = cor(x = data.long$experience, y = data.long$salary)
rm(list = c("nrobs", "experience", "salary"))
```
</details>   

```{r,echo=FALSE}
kable(data.long[1:5,], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```
Test for assumption of normality
```{r}
apply(X = data.long, MARGIN = 2, FUN = shapiro.test)
```

**Pearson correlation**
```{r}
cor.test(x=data.long$experience, y=data.long$salary, alternative = "two.sided", method = "pearson")
```

**Spearman correlation**
```{r}
cor.test(x=data.long$experience, y=data.long$salary, alternative = "two.sided", method = "spearman")
```
 ***
 
\newpage 

# Regression

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary> 
```{r}
# generate data
set.seed(05)
nrobs = 100
experience = rnorm(n = nrobs, mean = 15, sd = 3)
salary     = 10000 + ( 5 * experience ) + rnorm(n = nrobs, mean = 0, sd = 100)
data.long = data.frame(experience, salary)
# calculate correlation coefficient r
corr_coef = cor(x = data.long$experience, y = data.long$salary)
rm(list = c("nrobs", "experience", "salary"))
```
</details>   

```{r,echo=FALSE}
kable(data.long[1:5,], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Fit linear model
```{r}
linearModel = lm(formula = salary ~ experience, data = data.long)
```

Test for assumption of normality & equal variances
```{r}
# plot the residuals and look at the distribution around the 0-line and if the spread is equal over all x's.
plot(x=linearModel$fitted.values, y=linearModel$residuals, xlab = "predicted", ylab = "residuals"); abline(h=0)
shapiro.test(linearModel$residuals)
```
**Regression test**
```{r}
summary(linearModel)
```


***  
\newpage

# Logistic regression

***  
\newpage

#  One-way independent samples ANOVA
<details><summary><span style="color:dodgerBlue">Toon code om voorbeeld data te genereren </span></summary>
```{r, echo=TRUE}
set.seed(05)   # set seed
nrofconds = 3  # set number of conditions
nrofsubs  = 20 # set number of subjects
subj = as.factor(1:(nrofsubs*nrofconds))      # create array with subject IDs
condition = as.factor(rep(LETTERS[1:nrofconds],each=nrofsubs))   # create array with condition values
score = as.vector( replicate(
          nrofconds , rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1) ) 
        ) )                                     # create array with measurement values
data.long = data.frame(subj, condition, score);      # combine arrays into a data.frame
rm(list=c("subj", "condition","score","nrofconds","nrofsubs")) # delete unnecessary variables

```   
</details>  
   
```{r,echo=FALSE}
kable(data.long[c(1:2,21:22,41:42),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Test for assumption of normality
```{r}
by(data = data.long$score, INDICES = data.long$condition, FUN = shapiro.test)
```
Test for equality of variances
```{r}
leveneTest(y = data.long$score, group = data.long$condition)
```
**ANOVA*** with the aov method
```{r}
myModel = aov(formula = score ~ condition, data = data.long)
summary(myModel)
```
**ANOVA*** with the linear model method
```{r}
myModel = lm(formula = score ~ condition, data = data.long)
summary(myModel)
```

***  
\newpage

#  Factorial independent samples ANOVA

<details><summary><span style="color:dodgerBlue">Show code for data generation</span></summary>
```{r, echo=TRUE}
set.seed(01)   # set seed
nrofcondsf1 = 3  # set number of conditions for factor 1
nrofcondsf2 = 2  # set number of conditions for factor 2
nrofsubs    = nrofcondsf1*nrofcondsf2*30 # set number of subjects per condition
subj = as.factor(1:(nrofsubs))      # create array with subject IDs
# create array witht treatment conditions
treatment = as.factor(rep(LETTERS[1:nrofcondsf1],each=nrofsubs/nrofcondsf1))   
# create array with control / experimental
control   = as.factor(rep(c("control","experimental"),times=nrofsubs/nrofcondsf2))   
# create array with measurement values
score = as.vector( replicate(nrofcondsf1, replicate ( 
          nrofcondsf2 , rnorm(
            n = (nrofsubs/(nrofcondsf1*nrofcondsf2)), 
            mean = 0 , sd = sample(5,1) ) + sample(8,1)+10
        ) ) )                             
# combine arrays into a data.frame
data.long = data.frame(subj, score, treatment, control);      
# delete unnecessary arrays
rm(list=c("control","nrofcondsf1","nrofcondsf2","nrofsubs","score","subj","treatment"))

```   
</details>     

```{r,echo=FALSE}
kable(data.long[c(1:2,61:62,121:122),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Fit model with multiple prediction factors
```{r}
model = lm(formula = score ~ treatment+control, data = data.long)
```

Test for assumption of normality
```{r, eval=FALSE}
hist(model$residuals)
plot(x = fitted(model), y = residuals(model), xlab="predicted", ylab="residuals"); abline(h=0)
by(data = data.long$score, INDICES = paste(data.long$treatment, data.long$control), FUN = shapiro.test)
```
<details><summary><span style="color:dodgerBlue">Show residuals histogram</span></summary>
```{r, echo=FALSE}
hist(model$residuals)
```
</details>
<details><summary><span style="color:dodgerBlue">Show residuals vs predicted plot</span></summary>
```{r, echo=FALSE}
plot(x = fitted(model), y = residuals(model), xlab="predicted", ylab="residuals"); abline(h=0)
```
</details>
<details><summary><span style="color:dodgerBlue">Show output of Shapiro-Wilk test</span></summary>
```{r, echo=FALSE}
by(data = data.long$score, INDICES = paste(data.long$treatment, data.long$control), FUN = shapiro.test)
```
</details>

Test for equality of variances
```{r}
leveneTest(y = data.long$score, group = as.factor(paste(data.long$treatment, data.long$control)))
```

**ANOVA** with the linear model method
```{r}
model = lm(formula = score ~ treatment+control, data = data.long)
library(car)
Anova(mod = model, type = 'II')
summary(model)
```
\newpage 

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r, echo=TRUE}
set.seed(01)   # set seed
nrofcondsf1 = 3  # set number of conditions for factor 1
nrofcondsf2 = 2  # set number of conditions for factor 2
nrofsubs    = nrofcondsf1*nrofcondsf2*20 # set number of subjects per condition
subj = as.factor(1:(nrofsubs))      # create array with subject IDs
# create array witht treatment conditions
treatment = as.factor(rep(LETTERS[1:nrofcondsf1],each=nrofsubs/nrofcondsf1))   
# create array with control / experimental
control   = as.factor(rep(c("control","experimental"),times=nrofsubs/nrofcondsf2))   
# create array with measurement values
score = as.vector( replicate(nrofcondsf1, replicate ( 
          nrofcondsf2 , rchisq(
            n = (nrofsubs/(nrofcondsf1*nrofcondsf2)), 
            df = 3)  
        )+  sample(14,1)+10  ) )                             
# combine arrays into a data.frame
data.long = data.frame(subj, score, treatment, control);      
# delete unnecessary arrays
rm(list=c("control","nrofcondsf1","nrofcondsf2","nrofsubs","score","subj","treatment"))
```
</details>   


***  
\newpage  

#  One-way repeated measures ANOVA
<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r, echo=TRUE}
# Generate dataset
set.seed(01)   # set seed
nrofsubs  = 20 # set number of subjects
nrofconds = 3  # set number of conditions
subj = as.factor(rep(1:nrofsubs,nrofconds))      # create array with subject IDs
cond = as.factor(rep(LETTERS[1:nrofconds],each=nrofsubs))   # create array with condition values
score = as.vector( replicate(
          nrofconds , rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1) ) 
        ) )                                     # create array with measurement values
data.long = data.frame(subj, cond, score);      # combine arrays into a data.frame
rm(list=c("cond","nrofconds","nrofsubs","score","subj")) # delete arrays
```

```{r, echo=TRUE}
# Generate dataset
set.seed(01)   # set seed
nrofsubs  = 20 # set number of subjects
data.wide = data.frame(
  subj = as.factor(1:nrofsubs)   ,
  A =  rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1)) ,
  B =  rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1)) ,
  C =  rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1))
)
rm(list=c("nrofsubs")) # delete arrays
```

</details>  

```{r,echo=FALSE}
kable(data.long[c(1:2,21:22,41:42),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

fit linear model
```{r}
model = lm(formula = cbind(data.wide$A, data.wide$B, data.wide$C)~1)
```  
Test for assumption of normality
```{r}
hist(model$residuals)
shapiro.test(model$residuals)
```
Test for assumption of sphericity
```{r}
mauchly.test(model, X=~1)
```
ANOVA
```{r}
anova(model, X = ~1, test="Spherical")
```


***  

\newpage

#  Factorial repeated measures ANOVA

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r, echo=TRUE}
set.seed(02)  # set seed
nrofcondsf1 = 3  # set number of conditions for factor 1
nrofcondsf2 = 2  # set number of conditions for factor 2
nrofsubs    = 10 # set number of subjects 
# create array with subject IDs
subj = as.factor(rep(1:(nrofsubs),times=nrofcondsf1*nrofcondsf2))    
# create array witht treatment conditions
treatment = as.factor(rep(LETTERS[1:nrofcondsf1],each=nrofsubs*nrofcondsf2))   
# create array with control / experimental
control   = as.factor(
  rep(rep(c("control","experimental"),each=nrofsubs),times=nrofcondsf1))  
# create array with measurement values
score = as.vector( replicate(nrofcondsf1,        
                   replicate(nrofcondsf2,    
                   rnorm(n = (nrofsubs), mean = sample(14,1)+10 , sd = sample(5,1)        
                                   ) ) ) )                  
# combine arrays into a data.frame
data.long = data.frame(subj, score, treatment, control);      
# delete arrays
rm(list=c("control","nrofcondsf1","nrofcondsf2","nrofsubs", "score", "subj", "treatment"))
```
</details>   

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r, echo=TRUE}
set.seed(02)  # set seed
nrofsubs    = 20 # set number of subjects 

data.wide = data.frame(
  ID=1:nrofsubs,
  preA=rnorm(n=nrofsubs, mean = 09,sd = 3),
  preB=rnorm(n=nrofsubs, mean = 10,sd = 3),
  preC=rnorm(n=nrofsubs, mean = 11,sd = 3),
  postA=rnorm(n=nrofsubs, mean = 11,sd = 3),
  postB=rnorm(n=nrofsubs, mean = 15,sd = 3),
  postC=rnorm(n=nrofsubs, mean = 20,sd = 3)
)
rm(list="nrofsubs")
```
</details>   


```{r,echo=FALSE}
kable(data.long[1:4,], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Fit linear model to the data, not yet seperating both independent factors:
```{r}
model = lm(cbind(data.wide$preA,
                 data.wide$preB,
                 data.wide$preC,
                 data.wide$postA,
                 data.wide$postB,
                 data.wide$postC
                 )~1
           )
```
Make a factor matrix clarifying the conditions. Name the colomns as the factors.
Use the parts of the column names of the data.frame in the wide format as variable values in the matrix.
```{r}
factorMatrix = data.frame(
  time=as.factor(rep(c("pre","post"),each=3)),
  condition = as.factor(rep(c("A","B","C"),2))
)
```
fit linear model specifying the two independent variables using the factorMatrix
```{r}
library(car)
factorialAnova = Anova(mod=model, idata=factorMatrix, idesign= ~time*condition, type='III')
```
run the factorial ANOVA 
```{r}
summary(object = factorialAnova, multivariate=FALSE)
```


***  

\newpage

# Mixed-Design ANOVA

