---
title: "Overview statistic tests PB"
author: "Jonas van Nijnatten"
date: "21 november 2018"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
  pdf_document:
    fig_caption: yes
    number_sections: yes
    toc: yes
    toc_depth: '2'
header-includes:
- \usepackage{geometry}
- \geometry{a4paper, portrait, margin=.75in}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


contact: J.J.vanNijnatten@uva.nl

broncode: https://github.com/jonasvannijnatten/R_Data_Visualization  

***

\newpage

#   Software  
## Versies  
software versions used for this tutorial:  
 - `r R.Version()$version.string`  
 - car-package version: `r packageVersion("car")` (`r packageDate("car")`)
 
##  Installatie
Benodigde packages downloaden & installeren:
```{r, eval = FALSE, echo=TRUE, results='hide', message=FALSE}
install.packages(pkgs="kableExtra",   repos="https://www.freestatistics.org/cran/")
install.packages(pkgs="car",   repos="https://www.freestatistics.org/cran/")
```

Benodigde packages activeren:
```{r, eval=TRUE, message=FALSE}
library(package=kableExtra)
library(package=car)
```

***

\newpage

#  T-test

## Independent samples T-test
<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r}
# generate data
N = 40
data.long = data.frame(
  ID = 1:N,
  condition = rep(x = c("A","B"), each = N),
  score = c(rnorm(n = N, mean = 25, sd = 6.5), rnorm(n = N, mean = 35, sd = 6.5))
)
```
</details>

```{r,echo=FALSE}
kable(data.long[c(1:5,41:45),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Test for assumption of normality
```{r}
by(data = data.long$score, INDICES = data.long$condition, FUN = shapiro.test)
```
Test for equality of variances
```{r}
leveneTest(y = data.long$score, group = data.long$condition)
```
**T-test**
```{r}
t.test(formula = score ~ condition, data = data.long, paired=FALSE, alternative="two.sided", var.equal=TRUE)
```

## Paired samples T-test
<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r}
# generate data
N = 30
data.long = data.frame(
  ID = rep(1:N,2),
  condition = rep(x = c("A","B"), each = N),
  score = c(rnorm(n = N, mean = 25, sd = 6.5), rnorm(n = N, mean = 35, sd = 6.5))
)
```
</details>

```{r,echo=FALSE}
kable(data.long[c(1:5,31:35),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Test for assumption of normality
```{r}
# calculate difference scores
diffScore = data.long$score[data.long$condition=="A"] - data.long$score[data.long$condition=="B"]
shapiro.test(diffScore)
```

**T-test**
```{r}
t.test(formula = score ~ condition, data = data.long, paired=TRUE, alternative="two.sided", var.equal=TRUE)
```

***
\newpage

#  Correlation

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary> 
```{r}
# generate data
set.seed(05)
nrobs = 100
experience = rnorm(n = nrobs, mean = 15, sd = 3)
salary     = 10000 + ( 5 * experience ) + rnorm(n = nrobs, mean = 0, sd = 100)
data.long = data.frame(experience, salary)
# calculate correlation coefficient r
corr_coef = cor(x = data.long$experience, y = data.long$salary)
rm(list = c("nrobs", "experience", "salary"))
```
</details>   

```{r,echo=FALSE}
kable(data.long[1:5,], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```
Test for assumption of normality
```{r}
apply(X = data.long, MARGIN = 2, FUN = shapiro.test)
```

**Pearson correlation**
```{r}
cor.test(x=data.long$experience, y=data.long$salary, alternative = "two.sided", method = "pearson")
```

**Spearman correlation**
```{r}
cor.test(x=data.long$experience, y=data.long$salary, alternative = "two.sided", method = "spearman")
```
 ***
 
\newpage 

# Regression

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary> 
```{r}
# generate data
set.seed(05)
nrobs = 100
experience = rnorm(n = nrobs, mean = 15, sd = 3)
salary     = 10000 + ( 5 * experience ) + rnorm(n = nrobs, mean = 0, sd = 100)
data.long = data.frame(experience, salary)
# calculate correlation coefficient r
corr_coef = cor(x = data.long$experience, y = data.long$salary)
rm(list = c("nrobs", "experience", "salary"))
```
</details>   

```{r,echo=FALSE}
kable(data.long[1:5,], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Fit linear model
```{r}
linearModel = lm(formula = salary ~ experience, data = data.long)
```

Test for assumption of normality & equal variances
```{r}
# plot the residuals and look at the distribution around the 0-line and if the spread is equal over all x's.
plot(x=linearModel$fitted.values, y=linearModel$residuals, xlab = "predicted", ylab = "residuals"); abline(h=0)
shapiro.test(linearModel$residuals)
```
**Regression test**
```{r}
summary(linearModel)
```


***  
\newpage

#  One-way independent samples ANOVA
<details><summary><span style="color:dodgerBlue">Toon code om voorbeeld data te genereren </span></summary>
```{r, echo=TRUE}
set.seed(05)   # set seed
nrofconds = 3  # set number of conditions
nrofsubs  = 20 # set number of subjects
subj = as.factor(1:(nrofsubs*nrofconds))      # create array with subject IDs
condition = as.factor(rep(LETTERS[1:nrofconds],each=nrofsubs))   # create array with condition values
score = as.vector( replicate(
          nrofconds , rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1) ) 
        ) )                                     # create array with measurement values
data.long = data.frame(subj, condition, score);      # combine arrays into a data.frame
rm(list=c("subj", "condition","score","nrofconds","nrofsubs")) # delete unnecessary variables

```   
</details>  
   
```{r,echo=FALSE}
kable(data.long[c(1:2,21:22,41:42),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Test for assumption of normality
```{r}
by(data = data.long$score, INDICES = data.long$condition, FUN = shapiro.test)
```
Test for equality of variances
```{r}
leveneTest(y = data.long$score, group = data.long$condition)
```
**ANOVA*** with the aov method
```{r}
myModel = aov(formula = score ~ condition, data = data.long)
summary(myModel)
```
**ANOVA*** with the linear model method
```{r}
myModel = lm(formula = score ~ condition, data = data.long)
summary(myModel)
```

***  
\newpage

#  Factorial independent samples ANOVA

<details><summary><span style="color:dodgerBlue">Show code for data generation</span></summary>
```{r, echo=TRUE}
set.seed(01)   # set seed
nrofcondsf1 = 3  # set number of conditions for factor 1
nrofcondsf2 = 2  # set number of conditions for factor 2
nrofsubs    = nrofcondsf1*nrofcondsf2*30 # set number of subjects per condition
subj = as.factor(1:(nrofsubs))      # create array with subject IDs
# create array witht treatment conditions
treatment = as.factor(rep(LETTERS[1:nrofcondsf1],each=nrofsubs/nrofcondsf1))   
# create array with control / experimental
control   = as.factor(rep(c("control","experimental"),times=nrofsubs/nrofcondsf2))   
# create array with measurement values
score = as.vector( replicate(nrofcondsf1, replicate ( 
          nrofcondsf2 , rnorm(
            n = (nrofsubs/(nrofcondsf1*nrofcondsf2)), 
            mean = 0 , sd = sample(5,1) ) + sample(8,1)+10
        ) ) )                             
# combine arrays into a data.frame
data.long = data.frame(subj, score, treatment, control);      
# delete unnecessary arrays
rm(list=c("control","nrofcondsf1","nrofcondsf2","nrofsubs","score","subj","treatment"))

```   
</details>     

```{r,echo=FALSE}
kable(data.long[c(1:2,61:62,121:122),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

Fit model with multiple prediction factors
```{r}
model = lm(formula = score ~ treatment+control, data = data.long)
```

Test for assumption of normality
```{r, eval=FALSE}
hist(model$residuals)
plot(x = fitted(model), y = residuals(model), xlab="predicted", ylab="residuals"); abline(h=0)
by(data = data.long$score, INDICES = paste(data.long$treatment, data.long$control), FUN = shapiro.test)
```
<details><summary><span style="color:dodgerBlue">Show residuals histogram</span></summary>
```{r, echo=FALSE}
hist(model$residuals)
```
</details>
<details><summary><span style="color:dodgerBlue">Show residuals vs predicted plot</span></summary>
```{r, echo=FALSE}
plot(x = fitted(model), y = residuals(model), xlab="predicted", ylab="residuals"); abline(h=0)
```
</details>
<details><summary><span style="color:dodgerBlue">Show output of Shapiro-Wilk test</span></summary>
```{r, echo=FALSE}
by(data = data.long$score, INDICES = paste(data.long$treatment, data.long$control), FUN = shapiro.test)
```
</details>

Test for equality of variances
```{r}
leveneTest(y = data.long$score, group = as.factor(paste(data.long$treatment, data.long$control)))
```

**ANOVA** with the linear model method
```{r}
model = lm(formula = score ~ treatment+control, data = data.long)
library(car)
Anova(mod = model, type = 'II')
summary(model)
```
\newpage 

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r, echo=TRUE}
set.seed(01)   # set seed
nrofcondsf1 = 3  # set number of conditions for factor 1
nrofcondsf2 = 2  # set number of conditions for factor 2
nrofsubs    = nrofcondsf1*nrofcondsf2*20 # set number of subjects per condition
subj = as.factor(1:(nrofsubs))      # create array with subject IDs
# create array witht treatment conditions
treatment = as.factor(rep(LETTERS[1:nrofcondsf1],each=nrofsubs/nrofcondsf1))   
# create array with control / experimental
control   = as.factor(rep(c("control","experimental"),times=nrofsubs/nrofcondsf2))   
# create array with measurement values
score = as.vector( replicate(nrofcondsf1, replicate ( 
          nrofcondsf2 , rchisq(
            n = (nrofsubs/(nrofcondsf1*nrofcondsf2)), 
            df = 3)  
        )+  sample(14,1)+10  ) )                             
# combine arrays into a data.frame
data.long = data.frame(subj, score, treatment, control);      
# delete unnecessary arrays
rm(list=c("control","nrofcondsf1","nrofcondsf2","nrofsubs","score","subj","treatment"))
```
</details>   

```{r}

```


***  
\newpage  

#  One-way repeated measures ANOVA
<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r, echo=TRUE}
# Generate dataset
set.seed(01)   # set seed
nrofsubs  = 20 # set number of subjects
nrofconds = 3  # set number of conditions
subj = as.factor(rep(1:nrofsubs,nrofconds))      # create array with subject IDs
cond = as.factor(rep(LETTERS[1:nrofconds],each=nrofsubs))   # create array with condition values
score = as.vector( replicate(
          nrofconds , rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1) ) 
        ) )                                     # create array with measurement values
data.long = data.frame(subj, cond, score);      # combine arrays into a data.frame
rm(list=c("cond","nrofconds","nrofsubs","score","subj")) # delete arrays
```

```{r, echo=TRUE}
# Generate dataset
set.seed(01)   # set seed
nrofsubs  = 20 # set number of subjects
data.wide = data.frame(
  subj = as.factor(1:nrofsubs)   ,
  A =  rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1)) ,
  B =  rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1)) ,
  C =  rnorm(n = nrofsubs, mean = sample(8,1)+10 , sd = sample(5,1))
)
rm(list=c("nrofsubs")) # delete arrays
```

</details>  

```{r,echo=FALSE}
kable(data.long[c(1:2,21:22,41:42),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

fit linear model
```{r}
model = lm(formula = cbind(data.wide$A, data.wide$B, data.wide$C)~1)
```  
Test for assumption of normality
```{r}
hist(model$residuals)
shapiro.test(model$residuals)
```
Test for assumption of sphericity
```{r}
mauchly.test(model, X=~1)
```
ANOVA
```{r}
anova(model, X = ~1, test="Spherical")
```


***  

\newpage

#  Factorial repeated measures ANOVA

<details><summary><span style="color:dodgerBlue">Show code for data generation </span></summary>
```{r, echo=TRUE}
set.seed(02)  # set seed
nrofcondsf1 = 3  # set number of conditions for factor 1
nrofcondsf2 = 2  # set number of conditions for factor 2
nrofsubs    = 10 # set number of subjects 
# create array with subject IDs
subj = as.factor(rep(1:(nrofsubs),times=nrofcondsf1*nrofcondsf2))    
# create array witht treatment conditions
treatment = as.factor(rep(LETTERS[1:nrofcondsf1],each=nrofsubs*nrofcondsf2))   
# create array with control / experimental
control   = as.factor(
  rep(rep(c("control","experimental"),each=nrofsubs),times=nrofcondsf1))  
# create array with measurement values
score = as.vector( replicate(nrofcondsf1,        
                   replicate(nrofcondsf2,    
                   rnorm(n = (nrofsubs), mean = sample(14,1)+10 , sd = sample(5,1)        
                                   ) ) ) )                  
# combine arrays into a data.frame
data.long = data.frame(subj, score, treatment, control);      
# delete arrays
rm(list=c("control","nrofcondsf1","nrofcondsf2","nrofsubs", "score", "subj", "treatment"))
```
</details>   

```{r,echo=FALSE}
kable(data.long[c(1:2,11:12,21:22,31:32,41:42,52:52),], row.names = FALSE, caption = 'data.long') %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = FALSE, position = "left")
```

```{r}

```
  
***  



